---
output: html_document
editor_options: 
  chunk_output_type: console
  
---
Loan Data from Prosper by Yusuf Britton
========================================================

```{r global_options, include=TRUE, echo=FALSE}
#Used to suppress irrelavent warnings
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Main libs -------------------------

#install.packages('ggplot2')
library(ggplot2)

#install.packages('dplyr')
library(dplyr) #Used to chain functions, %>%

#Used to show multi reports with all the variables. Limit to the 15 used.
#install.packages('GGally')
library(GGally)

#Used to color different colors in multi reports
library(RColorBrewer)


```

```{r echo=FALSE, Load_the_Data}

# Data Sets -----------------

#read in csv file into dataframe
fullps <- read.csv('prosperLoanData-.csv')

#Selecting 15 variables for analysis
myvars <- c("Term", "LoanStatus", "BorrowerAPR", "ProsperRating..Alpha.", 
            "BorrowerState", "IsBorrowerHomeowner",  
            "CreditScoreRangeLower", "BankcardUtilization","DebtToIncomeRatio", 
            "StatedMonthlyIncome", "LoanOriginalAmount","MonthlyLoanPayment")

#subsetting dataframe
ps <- fullps[myvars]

#Change order of factor variable
rank = c("", "HR", " E", "D", "C", "B", "A", "AA")

ps$ProsperRating..Alpha. <- ordered(ps$ProsperRating..Alpha., 
                            levels = rank)


# Variables Manipulation
#New variables created, Annual income
ps$AnnualIncome = ps$StatedMonthlyIncome * 12

#variables for states choosen for multivariable analysis
psStates <- subset(ps, subset = 
               BorrowerState %in% c("CA", "FL", "GA", "IL", "NY", "OH", "TX"))

#Converting Term into a factor
ps$Term <- factor(ps$Term, levels=c(12,36,60), ordered=TRUE)

```


This is a data of over 113,000 borrowers that inquired for loans with Prosper.
This will look into some variables that may affect borrowers' APR or Prosper
grade. We will use 12 of the 81 variables in this analysis

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
#See shape of data
dim(ps)

#See names and types
#NOTE: interger are whole numbers, while numeric can be whole numbers and 
#decimals
str(ps)

#See statisical information
#NOTE: AmountDelinquent, DebtToIncomeRatio (1 should be highest), 
#StatedMonthlyIncome, MonthlyLoanPayment has outliers
summary(ps)
```

From looking at the statisical data, we have 113,937 observations 
over 12 variables.

One thing to note is that DebtToIncomeRatio (1 should be highest), 
StatedMonthlyIncome, and MonthlyLoanPayment have outliers.


## Terms

```{r echo=FALSE, Term_Plot}
ggplot(aes(x = Term), data = ps) + 
  geom_bar() 

```

From looking at the terms, it seems that 36 month is the most common terms for 
loans

## LoanStatus

```{r echo=FALSE, LoanStatus}
ggplot(aes(x = LoanStatus), data = ps) + 
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Most of the loans are still current or completed. There are a some that are
charged off or defaulted, while a small amount are pasted due or in final 
payment


## BorrowerAPR

```{r echo=FALSE, BorrowerAPR}
ggplot(aes(x = BorrowerAPR), data = ps) + 
  geom_histogram(binwidth = .001) + 
  scale_x_continuous(breaks = seq(.05,.4,.1), limits = c(.05,.4))

summary(ps$BorrowerAPR)

```

It seems that there is a spike APR around 36%. This will be one of the main 
variables to test what affect this variable. Average APR is about 21%


## ProsperRating (Alpha).

```{r echo=FALSE, ProsperRating..Alpha.}


ggplot(aes(x = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_histogram(stat="count") 

summary(ps$ProsperRating..Alpha.)
```

There are 29,084 out of the 113,937 that are blank. subsetting dataframe to 
only exclude blanks.From looking at the data, it seems that AA is the highest 
and HR is the lowest.

Also, most borrowers have a C rating

## BorrowerState

```{r echo=FALSE, fig.height=20, fig.width=20, BorrowerState}
ggplot(aes(x = BorrowerState), data = subset(ps, BorrowerState != "")) + 
  geom_histogram(stat="count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

There are 5,515 out of the 113,937 that are blank. subsetting dataframe to 
only exclude blanks.Borrowers are mostly from California. Florida, Illinois, 
New York, and Texas all follow, being a close second.
It may be from higher populations from bigger states or more populated cities.



## IsBorrowerHomeowner

```{r echo=FALSE, IsBorrowerHomeowner}
ggplot(aes(x = IsBorrowerHomeowner), data = ps) + 
  geom_histogram(stat="count") 

summary(ps$IsBorrowerHomeowner)

```

There are 57,478 home owners and 56,459 that don't own a home.


## CreditScoreRangeLower

```{r echo=FALSE, CreditScoreRangeLower}
ggplot(aes(x = CreditScoreRangeLower), data = ps) + 
  geom_histogram(binwidth = 20) +
  scale_x_continuous(breaks = seq(440,880,40), limits = c(440,880))

summary(ps$CreditScoreRangeLower)

table(ps$CreditScoreRangeLower)

```

For the lower credit score, it seems most borrowers have around 660-700

## BankcardUtilization

```{r echo=FALSE, BankcardUtilization}
ggplot(aes(x = BankcardUtilization), 
       data = subset(ps, BankcardUtilization <= 1)) + 
  geom_histogram(binwidth = .01) +
  scale_x_continuous(breaks = seq(0,1, .1), limits = c(0,1)) +
    ylim(0, 2250)

summary(ps$BankcardUtilization)

```
Per the variable definitions, Bank ultilaztion is a percentage. 
anything past 1 is an error.

Visually, more than of borrowers have more than 50% utilization

## DebtToIncomeRatio

```{r echo=FALSE, DebtToIncomeRatio}
ggplot(aes(x = DebtToIncomeRatio), data = subset(ps, DebtToIncomeRatio <= 1)) + 
  geom_histogram(binwidth = .01) 

summary(ps$DebtToIncomeRatio)

```
Per the variable definitions, DebtToIncomeRatio is a percentage. anything 
past 1 is an error.

Moat of the borrowers' debt to income ratio is 25% to 30% 

## Annual Income

```{r echo=FALSE, StatedMonthlyIncome}
#making annual income section
ps$AnnualIncome = ps$StatedMonthlyIncome * 12

ggplot(aes(x = StatedMonthlyIncome), data = ps) + 
  geom_histogram( binwidth = 100) +
  scale_x_continuous(breaks = seq(0,quantile(ps$StatedMonthlyIncome, .99, 
                                             na.rm = TRUE), 3000), 
                     limits = c(0,quantile(ps$StatedMonthlyIncome, .99, 
                                           na.rm = TRUE))) 

summary(ps$StatedMonthlyIncome)

ggplot(aes(x = AnnualIncome), data = ps) + 
  geom_histogram( binwidth = 2500) +
  scale_x_continuous(breaks = seq(0,quantile(ps$AnnualIncome, .99, na.rm = TRUE)
    ,25000), limits = c(0,quantile(ps$AnnualIncome, .99, na.rm = TRUE))) 


summary(ps$AnnualIncome)


```

Used 99% quantile to remove outliers.

Most borrowers monthly income is around $4,500 to $5,000, annual income 
ranging is around 40,000 to 60,000

## LoanOriginalAmount

```{r echo=FALSE, LoanOriginalAmount}
ggplot(aes(x = LoanOriginalAmount), data = ps) + 
  geom_histogram( binwidth = 100) +
  scale_x_continuous(breaks = seq(0,quantile(ps$LoanOriginalAmount, .99, 
                                             na.rm = TRUE),2000), 
                     limits = c(0,quantile(ps$LoanOriginalAmount, .99, 
                                           na.rm = TRUE))) 

summary(ps$LoanOriginalAmount)

```

Used 99% quantile to remove outliers.

Borrower's loan usual borrow around 4,000, 10,000, and 15,000. I wonder if the 
higher amounts are for homeowners?

## MonthlyLoanPayment

```{r echo=FALSE, MonthlyLoanPayment}
ggplot(aes(x = MonthlyLoanPayment), data = ps) + 
  geom_histogram(binwidth = 10) +
  scale_x_continuous(breaks = seq(0, quantile(ps$MonthlyLoanPayment, 
                                              .99, na.rm = TRUE),100), 
                     limits = c(0, quantile(ps$MonthlyLoanPayment, .99, 
                                            na.rm = TRUE))) 


summary(ps$MonthlyLoanPayment)

```

Used 99% quantile to remove outliers.

Most borrowers are paying around $150.00 in their monthly payments

### What is the structure of your dataset?
There are 113,937 diamonds in the dataset with 12 features (Term, Loan Status, 
Borrower APR, Prosper Rating, Borrower State, Borrower Homeownership, 
Credit Score, Bank Utilization, Debt to Income Ratio, Stated monthly income, 
Loan Original Amount, and Monthly Loan payment).

The variables prosper rating is a ordered factor variable with the 
following level.

(worst) ------> (best)
Prosper rating: HR, E, D, C, B, A, AA


Other observations:

- Average APR is 21%, with spikes around 36%
- Most borrowers have a C prosper rating
- Average credit score is about 685
- Average bank card utilization is 56%
- Debt to bank ratio is around 27%


### What is/are the main feature(s) of interest in your dataset?
The main feature in the data set is the Borrower APR and Prosper rating. 
I would like to determine what features affects the APR. 
I feel that the prosper grading, as well as other variables, 
affect borrow rate of interest.

### What other features in the dataset do you think will help support your \
I think Prosper rating, credit score, Bank Utilization, debt to bank ratio, 
Annual income, and home ownership may have an effect on the Borrower's APR

### Did you create any new variables from existing variables in the dataset?
I created the Annual income variable

### Of the features you investigated, were there any unusual distributions? \

I noticed the bank utilization and debt to income ratios have max values higher 
then 1.

From reading the variable dictionary, These values are ratio from 0 to 1.

I subsetted the file to exclude the outliers


# Bivariate Plots Section

```{r echo=FALSE, Prosper_rating_comparisons}
ggplot(aes(x = ProsperRating..Alpha., y = BorrowerAPR), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_jitter(alpha = .1) 

ggplot(aes(x = BorrowerState, y = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_jitter(alpha = .1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

From looking at the analysis, borrower's with a higher prosper rating tends 
to have a lower APR. 

From comparing prosper rating by state, I see concentrations in 
CA, FL, GA, IL, NY, OH, and TX. I'll look into these variables further


```{r echo=FALSE, credit_score_vs_APR}
ggplot(aes(x = CreditScoreRangeLower, y = BorrowerAPR), 
       data = subset(ps, CreditScoreRangeLower > 0)) + 
  geom_jitter(color = "#773F3F", alpha = .1)  +
  scale_x_continuous(breaks = seq(300,850,100), limits = c(300,850))


ggplot(aes(x = BorrowerState, y = CreditScoreRangeLower), 
       data = subset(ps, BorrowerState != "")) + 
  geom_jitter(color = "#773F3F", alpha = .1)  +
  scale_y_continuous(breaks = seq(300,850,100), limits = c(300,850)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

From research, The possible credit score is 300-850. set limits for this range. 

Visually, it seems that having a high credit score somewhat affect APR, but 
there is a huge APR range concentrate between 640 to 725. 

When comparing to states, the credits have greater rangings in states with
highly populated cities

There are probably other factors affecting this.

```{r echo=FALSE, Bank_Utilization_vs_APR}
ggplot(aes(x = BankcardUtilization, y = BorrowerAPR), 
       data = subset(ps, BankcardUtilization <= 1)) + 
  geom_jitter(color = "#773F3F", alpha = .05)


ggplot(aes(x = BankcardUtilization, y = CreditScoreRangeLower), 
       data = subset(ps, BankcardUtilization <= 1)) + 
  geom_jitter(color = "#773F3F", alpha = .05)

```

visually, it doesn't seem like bank utilization has a major affect on APR. 
One thing to note is that regardless of what utilization is, 
there is a concentration around 36% APR.

From comparing bank utilization to credit score, it seems that borrowers with 
lower bank utilization tends to
have higher credit scores

```{r echo=FALSE, debt_to_bank_ratio_vs_APR}
ggplot(aes(x = DebtToIncomeRatio, y = BorrowerAPR), 
       data = subset(ps, DebtToIncomeRatio <= 1)) + 
  geom_jitter(color = "#773F3F", alpha = .05)

ggplot(aes(x =  IsBorrowerHomeowner, y = DebtToIncomeRatio), 
       data = subset(ps, DebtToIncomeRatio <= 1)) + 
  geom_jitter(color = "#773F3F", alpha = .05)

```

visually, it doesn't seem like debt to bank ratio has a major affect on APR. 
One thing to note is that regardless of what utilization is, there is a 
concentration around 36% APR

From comparing home ownership to debt to bank, there doesn't seem to be much 
of a difference 

```{r echo=FALSE, Annual_income_vs_APR}
ggplot(aes(x = AnnualIncome, y = BorrowerAPR), data = ps) + 
  geom_jitter(color = "#773F3F", alpha = .5) +
  scale_x_continuous(breaks = seq(0,quantile(ps$LoanOriginalAmount, .99, 
                                             na.rm = TRUE),10000), 
                     limits = c(0,quantile(ps$LoanOriginalAmount, .99, 
                                           na.rm = TRUE)))
```

visually, it doesn't seem like debt to bank ratio has a major affect on APR. 
One thing to note is that regardless of what utilization is, there is a 
concentration around 36% APR

```{r echo=FALSE, IsBorrowerHomeowner_vs_APR}
ggplot(aes(x = IsBorrowerHomeowner, y = BorrowerAPR), data = ps) + 
  geom_jitter(color = "#773F3F", alpha = .01) 

```

visually, it doesn't seem like owning a home has a major affect on APR. 
One thing to note is that regardless if a borrow owns home, there is a 
concentration around 36% APR

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \

APR correalates to prosper rating and credit score, which looks like it's 
batching ranging of credit score

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
What I found interesting was the comparison of states to the prosper data. 
I will look in how CA, FL, GA, IL, NY, OH, and TX compare to one another in 
the multivariate analysis



### What was the strongest relationship you found?
Between APR and Prosper Grade

# Multivariate Plots Section

```{r echo=FALSE, ProsperRating..Alpha._multivariate}

ggplot(aes(x =  BorrowerState, y = BorrowerAPR, 
           color = ProsperRating..Alpha.), 
       data = subset(psStates, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = 1, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  ggtitle('APR by Prosper score and State')

```

Each state follow the similar breakdowns between APR and prosper rating. CA 
and TX has higher concentration around have less than 10% APR. In IL, there are 
some HR rating that were able to get better rates.



```{r echo=FALSE, CreditScoreRangeLower_multivariate}

ggplot(aes(x = CreditScoreRangeLower, y = BorrowerAPR, 
           color = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = 1, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  ggtitle('APR by credit score and Prosper score')



```

From looking at the credit scores, there are some HR ratings that got better 
rate than other. 

Otherwise, the relationship looks like borrowers for better credit scores have 
better APR and prosper rating.

```{r echo=FALSE, BankcardUtilization_multivariate}

ggplot(aes(x = BankcardUtilization, y = BorrowerAPR, 
           color = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = .5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  scale_x_continuous(limits = c(0.0, 1), breaks = c(0.25, 0.5, .75, 1)) + 
  ggtitle('APR by credit card utilization and Prosper score')


```

As bank card utilization goes up, you see less AA rating and more HR ratings.


```{r echo=FALSE, DebtToIncomeRatio_multivariate}

ggplot(aes(x = DebtToIncomeRatio, y = BorrowerAPR, 
           color = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = .5, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  scale_x_continuous(limits = c(0.0, 1), breaks = c(0.25, 0.5, .75, 1)) + 
  ggtitle('APR by debt to income ratio and Prosper score')



```

For the rating, most AA are in the 0 to 40 percent range. HR is almost as high 
as 100%. All the other ratings are around 60%

```{r echo=FALSE, AnnualIncome_multivariate}

ggplot(aes(x = AnnualIncome, y = BorrowerAPR, color = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = .05, size = 2, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", 
                                          reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  scale_x_continuous(limits = c(0,quantile(ps$AnnualIncome, .99, na.rm = TRUE)), 
                     breaks = seq(0,quantile(ps$AnnualIncome, .99, 
                                             na.rm = TRUE),50000)) + 
  ggtitle('APR by Annual Income and Prosper score')

```


```{r echo=FALSE, IsBorrowerHomeowner_multivariate}

ggplot(aes(x = IsBorrowerHomeowner, y = BorrowerAPR, 
           color = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = .05, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  ggtitle('APR by Annual Income and Prosper score')

```

Interestly, there are more homeowners that have AA ratings.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \

From comparing variables to Prosper rating and APR, each variable followed a 
pretty similar relationship from what was explored in the bivariate plots

### Were there any interesting or surprising interactions between features?

I found it surprising the homeowners had higher prosper ratings than 
non-homeowners


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One, include=TRUE}
ggplot(aes(x = BorrowerState, y = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_jitter(alpha = .1) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle('Borrower state vs Prosper Rating')

psStates %>%
  group_by(BorrowerState) %>%
  summarise (n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  arrange(desc(freq))

ps %>%
  group_by(BorrowerState) %>%
  summarise (n = n()) %>%
  mutate(freq = n / sum(n)) %>%
  filter(BorrowerState %in% c("CA", "FL", "GA", "IL", "NY", "OH", "TX")) %>%
  arrange(desc(freq))

```


### Description One
What I found interesting about this is the intensity of ratings in different 
states.
it could be that the states with the higher concentration were because 
some they contain some of America's biggest cities.

In this set, CA makes up 29% of the 6 states chosen in the analysis and 13% of 
the population

### Plot Two
```{r echo=FALSE, Plot_Two, include=TRUE}
ggplot(aes(x = CreditScoreRangeLower, y = BorrowerAPR), 
       data = subset(ps, CreditScoreRangeLower > 0)) + 
  geom_jitter(color = "#773F3F", alpha = .1)  +
  scale_x_continuous(breaks = seq(300,850,100), limits = c(300,850)) +  
  ggtitle('Credit score vs APR')


summary(ps$CreditScoreRangeLower)
```

### Description Two
Though some of the borrowers had perfect credit scores, they still had a 
high APR interest rating.

Most of the borrowers credit scores ranged from 660 to 720.

### Plot Three
```{r echo=FALSE, Plot_Three, include=TRUE}
ggplot(aes(x = IsBorrowerHomeowner, y = BorrowerAPR, 
           color = ProsperRating..Alpha.), 
       data = subset(ps, ProsperRating..Alpha. != "")) + 
  geom_point(alpha = .05, size = 1, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = "Prosper rating", 
                                          reverse = T,
                                          override.aes = list(alpha = 1, 
                                                              size = 2))) +  
  ggtitle('APR by Annual Income and Prosper score')

ps %>%
  group_by(IsBorrowerHomeowner, ProsperRating..Alpha.) %>%
  summarise (n = n()) %>%
  mutate(Percentage = n / sum(n)) %>%
  filter(ProsperRating..Alpha. == "AA") %>%
  arrange(desc(Percentage))
```



### Description Three
Home owners generally had more AA ratings. It could be that homeownership plays
a big part on getting a better rating.

There are 3,847 homeowners in the AA rating, and 1,525 non-homeowners in this
rating


------

# Reflection


From doing the analysis, I was able to find that APR follow pretty closely with 
the prosper rating.Some of the struggle I had was trying to find the 
relationship, only to realize that some of the information isn't complete. 
For example, I would have loved to look at rating by occupation, but quickly 
saw that the "Professional" value made up most of the variable. I found that 
there was alot of missing information.

I was surprised about the impact homeownership have on securing a AA rating 
and having a decent APR.

I would like to look back at this data set by acquiring a more complete data 
set that would show a better breakdown of occupation to do a further analysis,
as well as retrieving city data in states that has the highest borrowers


